Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2024-07-08 paper/analysis.tex
pull feedback from Daniel

\paragraph{Nyström estimator} Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}

@@

@@

    Setting $\gamma = e$ and choosing $k = \lfloor \log(\delta^{-1}) / 2 \rfloor$,
    %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
    %we get that for all $\delta \in (0, 1)$
    the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil (2^3 e \varepsilon^{-1} \log(\delta^{-1}))^2 \rceil$.
    %where $\lceil x \rceil$ is the ceiling function which returns the smallest integer larger than or equal to $x$.
    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
    %\end{equation}
\end{proof}
\color{black}

%\begin{remark}
%    If additionally $\mtx{B}(t)$ is nonzero and symmetric positive semi-definite for all $t \in [a, b]$, then it follows from the proof of \refthm{thm:hutchinson} that with probability at least $1 - \delta$
%    \begin{equation}
%        \int_{a}^{b} \frac{| \Trace(\mtx{B}(t)) - \Hutch{\mtx{\Psi}}(\mtx{B}(t)) |}{\Trace(\mtx{B}(t))} ~\mathrm{d}t < \varepsilon,
%    \end{equation}
%    if $n_{\mtx{\Psi}} \geq 2^{10} \varepsilon^{-2} \log(\delta^{-1})^2 (\int_{a}^{b} \mu(t)^{\sfrac{1}{2}}~\mathrm{d}t)^2$ with $\mu(t) = \lVert \mtx{B}(t) \rVert _2 / \Trace(\mtx{B}(t))$. Compared to the constant matrix equivalent \cite[Remark 2]{cortinovis-2022-randomized-trace} we mainly observe a larger constant and an additional factor of $\log(\delta^{-1})$.
%\end{remark}

Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the result of~\refthm{thm:hutchinson} requires one to choose the number of queries $n_{\mtx{\Psi}}$ by a factor $\log(\delta^{-1})$ larger to yield the same guarantee.

@@

For symmetric positive semi-definite matrices whose singular values decay rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be significantly better than the Girard-Hutchinson estimate. This is reflected in the following theorem.

\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
    Suppose $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ is continuous and positive semi-definite in $t \in [a, b]$. Let $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$ be two integers. Then for all $\gamma \geq 1$ with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$
    \begin{equation}
        \int_{a}^{b} \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right| ~\mathrm{d}t
        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(\mtx{B}(t)) ~\mathrm{d}t.
    \end{equation}
\end{theorem}

\begin{proof}
    Since $\mtx{B}(t)$ is positive semi-definite for all $t$, then by \cite[Lemma 2.1]{frangella-2023-randomized-nystrom}, $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is also positive semi-definite for all $t$. Therefore, for fixed $t \in [a, b]$
    \begin{equation}
        \left| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
        = \left| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \right|
        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast},
    \end{equation}
    where $\lVert \cdot \rVert _{\ast}$ represents the nuclear norm.  From the proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms} it follows that
    \begin{equation}
        %\lVert f(\mtx{A}, t) - \widehat{f}(\mtx{A}, t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _2^2 \leq \lVert (\mtx{I}_n - \mtx{\Pi}_{f(\mtx{A}, t)^{\sfrac{1}{2}} \mtx{\Omega}}) f(\mtx{A}, t)^{\sfrac{1}{2}} \rVert _F^2
        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2.
    \end{equation}
    Hence, by \cite[Theorem 9]{kressner-2023-randomized-lowrank}, with $n_{\mtx{\Omega}} = p + r$, the result follows by noting $\sigma_i(\mtx{B}(t)^{\sfrac{1}{2}})^{2} = \sigma_i(\mtx{B}(t))$.

Compared to the constant matrix case \cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, our bound also scales with the best rank-$r$ approximation error in the nuclear norm, albeit with a factor approximately proportional to $r$, which cannot be compensated for with large enough oversampling.

The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly, for example $\mtx{B}(t) \equiv \mtx{I}_n$.

>>>>

\paragraph{Nyström estimator}

Write:
\begin{itemize}
 \item fast singular value decay, intuitively sufficient to consider a low-rank approximation
 \item motivated by our application we only consider symmetric positive semi-definite case here (smoothing kernels are positive) for which Nystroem approximation is the right choice
\end{itemize}

\textcolor{red}{Moved from below:
For symmetric positive semi-definite matrices whose singular values decay rapidly, the Nyström estimate \refequ{equ:nystrom-trace-estimator} can be significantly better than the Girard-Hutchinson estimate. This is reflected in the following theorem.}

Alternatively, the trace of a symmetric matrix whose singular values decay quickly can be approximated well by using a Gaussian sketching matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ to form the Nyström approximation \cite{gittens-2013-revisiting-nystrom}

@@

\textcolor{red}{Moved from below: The downside of the Nyström estimator is that it will -- in general -- not work well enough on matrices whose singular values do not decay quickly, for example $\mtx{B}(t) \equiv \mtx{I}_n$.}

@@

    Setting $\gamma = e$ and choosing $k = \lfloor \log(\delta^{-1}) / 2 \rfloor$,
    %where $\lfloor x \rfloor$ is the floor function which returns the largest integer smaller than or equal to %$x$,
    %we get that for all $\delta \in (0, 1)$
    the second part of the theorem follows when taking $n_{\mtx{\Psi}} = \lceil (2^3 e \varepsilon^{-1} \log(\delta^{-1}))^2 \rceil$.
    %where $\lceil x \rceil$ is the ceiling function which returns the smallest integer larger than or equal to $x$.
    %Consequently, with probability $1 - \delta$ we have
    %\begin{equation}
    %    \int_{a}^{b} \left| \frac{1}{n_{\mtx{\Psi}}} \sum_{i=1}^{n_{\mtx{\Psi}}} \vct{\psi}_i^{\top} \mtx{B}(t) \vct{\psi}_i - \Trace(\mtx{B}(t)) \right| ~ \mathrm{d}t \leq 32 e^{\sfrac{1}{2}} \frac{\log(\delta^{-1})}{\sqrt{n_{\mtx{\Psi}}}} \int_{a}^{b} \lVert \mtx{B}(t) \rVert _F~\mathrm{d}t
    %\end{equation}
\end{proof}

Compared to the analogous result for the constant case~\cite[Lemma 2.1]{meyer-2021-hutch-optimal}, the number of queries $n_{\mtx{\Psi}}$ needed by~\refthm{thm:hutchinson} is roughly a factor $\log(\delta^{-1})$ larger.

@@

The following result confirms the inituition that the Nystr\"om approximation is very well suited in the presence of decaying singular values: If the nuclear norm error to the best rank-$r$ approximation to $B(t)$ is bounded by $\varepsilon$ for every $t$ then the trace of the Nystr\"om approximation (with a bit of oversampling) results, with high probability, in an error that is only modestly larger than $\varepsilon$.

\begin{theorem}{Nyström estimator for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be symmetric positive semi-definite and continuous for $t \in [a, b]$. For integers $r \geq 2$ and $n_{\mtx{\Omega}} \geq r + 4$, consider the Nystr\"om approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ defined in~\eqref{equ:nystrom-approximation} with an $n\times n_{\mtx{\Omega}}$ Gaussian random matrix $\mtx{\Omega}$. Then for any $\gamma \geq 1$ the bound 
    \[
        \int_{a}^{b} \big| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big| ~\mathrm{d}t
        \leq \gamma^2 (1 + r) \int_{a}^{b} \sum_{i = r+1}^{n} \sigma_i(\mtx{B}(t)) ~\mathrm{d}t.
    \]
    holds with probability at least $1 - \gamma^{-(n_{\mtx{\Omega}} - r)}$.
\end{theorem}
\begin{proof}
Using that $\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ is positive semi-definite (see, e.g.,~\cite[Lemma 2.1]{frangella-2023-randomized-nystrom}) we obtain
    \begin{align*}
        \big| \Trace(\mtx{B}(t)) - \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big|
        & = \big| \Trace(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)) \big|
        = \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _{\ast} \\
        & = \lVert (\mtx{I}_n - \mtx{\Pi}_{\mtx{B}(t)^{\sfrac{1}{2}} \mtx{\Omega}}) \mtx{B}(t)^{\sfrac{1}{2}} \rVert _F^2,
    \end{align*}
    where $\lVert \cdot \rVert _{\ast}$ denotes the nuclear norm and we used a \textcolor{red}{result from the proof of \cite[Corollary 8.2]{tropp-2023-randomized-algorithms} ??? Is this really the original source. Have a look whether David P mentions this somewhere? Also, this may be contained in the original Gittens Mahoney paper but I am not sure ???} in the last equality. This allows us to apply \textcolor{red}{Theorem 9 ??? be careful that the theorem numbering may be different when the paper is published in NLAA ???} from \cite{kressner-2023-randomized-lowrank}, which immediately implies the statement claimed by the theorem.
\end{proof}

Compared to the constant case~\cite[Theorem 8.1]{tropp-2023-randomized-algorithms}, the bound of Theorem~\ref{thm:nystrom} features an additional factor $1+r$, which cannot be compensated with oversampling.

### 2024-07-08 paper/paper.tex
Updates from Overleaf
pull feedback from Daniel

@@

\textcolor{red}{TODO for Fabio:
\begin{itemize}
 \item Capitalize references to algorithms, lemma etc (that is, Lemma 3.2 instead of lemma 3.2 [Lin17, Algorithm 1] instead of [Lin17, algorithm 1]), etc. \textcolor{green}{Done.}
 \item Add full stops after paragraph titles. \textcolor{green}{Done.}
 \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style. The paper needs to get strictly below 30 pages prior to submission. \textcolor{green}{Journal abbreviations, max 2 authors, omit superfluous fields $\to$ 3 pages.}
 \item Please include Acknowledgements as indicated in the comment. \textcolor{green}{Done.}
\end{itemize}
}
\color{blue}
\input{intro.tex}
\input{analysis.tex}
\input{spectraldensity.tex}
\input{numresults.tex}

>>>>

% to speed up compilation speed a bit while drafting
\pdfcompresslevel=0
\pdfobjcompresslevel=0

@@

\textcolor{red}{TODO for Fabio:
\begin{itemize}
 \item It is not great that the references occupy so much space (more than 4 pages); consider using a more condensed style. The paper needs to get strictly below 30 pages prior to submission. \textcolor{green}{Journal abbreviations, max 2 authors, omit superfluous fields $\to$ 3 pages.}
 Journal abbreviations and removing superfluous fields is good. max 2 authors is not coming in math, please reverse this. The text style is a bit too generous but no need to change this now.
\end{itemize}
}
\color{blue}
\input{intro.tex}
\input{analysis.tex}
\input{spectraldensity.tex}
%\input{numresults.tex}
